---
title: "Hierarchy of time"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hierarchy of time}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  # collapse = TRUE,
  # comment = "#>",
  message = FALSE, 
  warning = FALSE, 
  # eval = identical(Sys.getenv("NOT_CRAN"), "true"),
  dev = "png",
  dpi = 120,
  fig.asp = 0.4,
  fig.width = 10,
  out.width = "100%",
  fig.align = "center"
)
options(crayon.enabled = TRUE)
options(pillar.bold = TRUE, pillar.subtle_num = TRUE)

# eulachon <-
#   readRDS('../fish/2022-10-02-trawl.rds') |>
#   dplyr::filter(grepl('Thaleichthys', species)) |>
#   dplyr::select(-trawl_id, -lat, -lng, -species, -flourescence, -turb, -ld, -rd,
#                 -date, -cpue_kg, -cpue_count) |>
#   dplyr::relocate(year, .before = 1)
# usethis::use_data(eulachon, overwrite = TRUE)
```

## Overview

This vignette presents a hierarchical Bayesian workflow for modelling temporal (year-to-year) changes n the distribution of discrete **species abundance**, $y$, along an **environmental gradient**, $x$, using the **modskurt** function to predict mean abundance, $\mu$. The model is specified as 

...

## Load packages and stan files

This section assumes you've already got `modskurt` installed and running, if not, jump over to this packages' [home page](../index.html).

```{r load-compile}
# compile_stanmodels only runs when needed, so it doesn't hurt to keep these
# lines together in any analysis
library(modskurt)
compile_stanmodels()

# everyone loves a good graph
library(ggplot2)
library(patchwork)
theme_set(theme_classic())

# for reproducibility
set.seed(123456)
```

## Case study - Eulachon vs. sea temperature over time

Use Northwest Pacific groundfish trawl survey data for *Thaleichthys pacificus* (eulachon, or candlefish) over the ocean temperature gradient - [more info](https://www.fisheries.noaa.gov/west-coast/science-data/us-west-coast-groundfish-bottom-trawl-survey). Key variables are:

- `total_count` the number of individual redstripes caught during each trawl.
- `temp` average water temperature recorded at the net during the trawl.
- `area_swept` total area (ha) of water swept during the trawl (trawl length * net mouth diameter), this is the sampling **effort** used to observe `total_count`.
- `year` year in which the data was collected.

```{r data}
data(eulachon)
eulachon
```

## A. Specify initial model

We'll start with the simplest model that ignores time for now,

$$
\begin{aligned}
y_n &\sim \text{NB}\left[\text{area_swept} \cdot \text{Modskurt}(x_n), \kappa\right] \\
\end{aligned}
$$

```{r}
# specify simple model for latest year
spec <-
  mskt_spec(data = eulachon[eulachon$year == max(eulachon$year),],
            y = c('Eulachon (CPUE)' = 'total_count'),
            x = c('SST (°C)' = 'temp'),
            effort = c('Area swept (ha)' = 'area_swept'),
            dist = 'nb',
            shape = 'rdp',
            subset_prop = 0.3)
```

```{r model-spec}
str(spec)
```

## B. Verify initial model specification

### Prior probabilities of parameter values

```{r check-prior-dens, fig.asp=0.8, fig.cap=""}
# check_prior_dens(spec)
```

### Prior predictions for the `modskurt` mean

```{r check-prior-mskt, fig.cap="TODO: add legend for alpha!"}
# check_prior_mskt(spec)
```

### Prior predictions for summary statistics of `y`

```{r check-prior-dist, fig.asp=0.3}
check_prior_dist(spec)
```

*OPTIONAL*: refine model spec further

## C. Fit subset model

```{r fit-subset, results='hide'}
fit_subset <-
  mskt_fit(spec,
           # use the subset only
           use_subset = TRUE,
           iter_warmup = 200,
           iter_sampling = 100,
           chains = 6,
           parallel_chains = 6,
           # for testing
           show_messages = TRUE, show_exceptions = TRUE)
```

```{r fit-subset-check}
check_computation(fit_subset)
```

*OPTIONAL*: refine model spec or fit parameters

### Check subset model

```{r check-post-dens-subset, fig.asp=0.8}
# visual display of chain mixing and prior-data conflicts
check_post_dens(fit_subset, by_chain = TRUE)
```

```{r check-post-calibration-subset}
# uses test set for discrete pit
check_post_calibration(fit_subset)
```

*OPTIONAL*: refine model spec or fit parameters

## D. Fit full model

```{r fit-full, results='hide'}
fit_full <-
  mskt_fit(spec,
           chains = 4,
           parallel_chains = 4,
           # for testing
           show_messages = TRUE, show_exceptions = TRUE)
```

```{r fit-full-check}
check_computation(fit_full)
```

*OPTIONAL*: refine model spec or fit parameters

### Check full model

```{r check-post-influencers-full}
check_post_influencers(fit_full)
```

## E. Use the model

### Plot summaries of the abundance distribution

```{r use-full-fit-dist}
# plot the distribution of abundance along the gradient
abundance_dist(fit_full, summaries = c('mean', 'median', 'q90')) +
  scale_y_sqrt()
```

### Calculate ranges of x for different percentages of abundance measures

```{r use-full-fit-range-plot}
# compose in row using patchwork
# (abundance_range(fit_full,
#                  capture_pct = 50,
#                  using_range = 'HAZ',
#                  based_on = 'mean',
#                  plotted = TRUE,
#                  range_colour = 'red') +
#    labs(subtitle = '50% of highest mean counts')) + 
#   (abundance_range(fit_full,
#                    capture_pct = 90,
#                    using_range = 'ADL',
#                    based_on = 'mean',
#                    # specify left or right side limit
#                    region = 'left',
#                    plotted = TRUE,
#                    range_colour = 'blue') +
#      labs(y = NULL, subtitle = '90% limit of mean count density')) +
#   plot_layout(guides = 'collect')
```

## Summary
