---
title: "Linked zero-inflation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linked zero-inflation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
resource_files:
  - '../man/figures/workflow.JPG'
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  # collapse = TRUE,
  # comment = "#>",
  message = FALSE, 
  warning = FALSE, 
  # eval = identical(Sys.getenv("NOT_CRAN"), "true"),
  dev = "png",
  dpi = 120,
  fig.asp = 0.4,
  fig.width = 10,
  out.width = "100%",
  fig.align = "center"
)
options(crayon.enabled = TRUE)
options(pillar.bold = TRUE, pillar.subtle_num = TRUE)

# redstripe <-
#   readRDS('../fish/2022-10-02-trawl.rds') |>
#   dplyr::filter(species == 'Sebastes proriger') |>
#   dplyr::select(-trawl_id, -lat, -lng, -species, -flourescence, -turb, -ld, -rd, -year,
#                 -cpue_kg, -cpue_count) |>
#   dplyr::relocate(date, .before = 1)
# usethis::use_data(redstripe, overwrite = TRUE)
```

## Overview

The [Getting started article](./getting-started.html) showed how to model the negative binomial distribution of species abundance along an environmental gradient using the `modskurt` mean function. We've found that in the vast majority of species-environment relationships, abundance can be reasonably characterised by a negative binomial distribution. However, for very rare and hard to detect species, there can be zero counts in "excess" of what the negative binomial distribution *should* predict. We say *should*, because mathematically, the negative binomial distribution can predict extremely high proportions of zeros, although to achieve this we require extremely high dispersion parameter values. Which is quite restricting in practice, the mechanisms that can cause high zero proportions may not also produce extremely disperse data.

To increase zero-rates, while maintaining realistic dispersion of abundance, a zero-inflated negative binomial can be used where a point probability mass is added at zero. This can work well in simple models, but in generalised models like these `modskurt` ones, there is often a relationship between mean abundance and zero rates in different environmental conditions. For example, mean abundances and zero rates are usually quite different in conditions where the species is commonly found and easy to detect, versus intolerable conditions that are hard to sample the species in. For this reason, the `modskurt` model utilises a **z**ero-**i**inflated **n**egative **b**inomial with excess-zero probability **l**inked (ZINBL) to mean abundances through a logistic regression. The formula for this distribution will be covered later, as we work through another case study.

## Load packages and stan files

```{r load-compile}
library(modskurt)
compile_stanmodels()

# everyone loves a good graph
library(ggplot2)
library(patchwork)
theme_set(theme_classic())

# for reproducibility
set.seed(123456)
```

## Case study - *Sebastes proriger* vs. water temperature

`ZINBL` distributions of abundance are typically most fruitful when proportions of zeros are very high, like $>0.95$ high. A great example of this is the Northwest Pacific groundfish trawl survey data for *Sebastes proriger* (redstripe rockfish) over the ocean temperature gradient - [ more info](https://www.fisheries.noaa.gov/west-coast/science-data/us-west-coast-groundfish-bottom-trawl-survey). Key variables are:

- `total_count` the number of individual redstripes caught during each trawl.
- `temp` average water temperature recorded at the net during the trawl.
- `area_swept` total area (ha) of water swept during the trawl (trawl length * net mouth diameter), this is the sampling **effort** used to observe `total_count`.

```{r data}
data(redstripe)
redstripe
```

## A. Specify initial model

The `ZINBL` distribution is defined as a mixture of the negative binomial distribution and a bernoulli process with mixing proportion, $\pi$ modelled as a decreasing logistic function of `modskurt` mean abundance, $\mu$,

$$
\begin{aligned}
y_n &\sim \text{ZINBL}\left[\mu_n = \text{Modskurt}(x_n), \kappa, \pi_n\right] \\
&= \pi_n \cdot (y_n == 0) + (1 - \pi_n) \cdot \text{NB}(\mu_n, \phi) \\
\text{logit}^{-1}~\pi_n &= \gamma_0 - \gamma_1 \cdot \mu_n
\end{aligned}
$$

with new parameters:

- $\gamma_0 \in \mathbb{R}$: log-odds intercept for the probability of observing an excess-zero,
- $\gamma_1 > 0$: describes the rate at which the excess-zero probability decreases as the average abundance increases. As $\gamma_0 \to 0$ or $\gamma_1 \to \infty$, $\pi \to 0$ and the ZINBL reduces to the NB distribution.

Muck around with parameters [here](https://salt-ecology.shinyapps.io/nb-zinbl-prior-specs/).

As with the negative binomial model, the `modskurt` package suggests some default priors that *should* just work 🤞🤞:

$$
\begin{align}
\kappa &\sim \text{Exponential}(0.5) \\
\gamma_0 &\sim \text{Normal}(3.0, 1.5) \\
\gamma_1 &\sim \text{HalfNormal}(5.0)
\end{align}
$$

Note, now that we have sampling effort, we can internally standardise counts to catch per unit effort (CPUE), $y / \text{area_swept}$, in the model using

$$
y_n \sim \text{ZINBL}\left[\text{area_swept} \cdot \mu_n, \kappa, \pi_n\right]
$$

Specifying the model in `R`:

```{r model}
spec <-
  mskt_spec(data = redstripe,
            # total_count is being modelled as CPUE by including effort
            y = c('Sebastes proriger (CPUE)' = 'total_count'),
            x = c('SST (°C)' = 'temp'),
            effort = c('Area swept (ha)' = 'area_swept'),
            dist = 'zinbl',
            shape = 'rdp',
            subset_prop = 0.3)
```

A quick look at the model spec for the full dataset shows that the zero proportion is about $2252 / 2330 = 0.97$:

```{r model-spec}
str(spec$full())
```

## B. Verify initial model specification

This is very much the same as in [getting started](./getting-started.html), except now we can have a look at the new linked zero inflation parameters, and the excess-zero probabilities they predict.

### Prior probabilities of parameter values

```{r check-prior-dens, fig.asp=0.8}
check_prior_dens(spec)
```

### Prior predictions for the `modskurt` mean

Note above that values for `m` are by default constrained within the range of `x` that had at least one non-zero abundance, `y`, recorded. The effect of this is quite visible when looking at prior predictions for mean abundance:

```{r check-prior-mskt, fig.cap="Line opacity shows the prior probability of jointly observing the parameters that predict that shape"}
check_prior_mskt(spec)
```

### Prior predictions for summary statistics of $y$

Dispersion is a little high again?

```{r check-prior-dist, fig.asp=0.3}
check_prior_dist(spec)
```

### Prior predictions for probability of excess zero $\pi$

This is a new test, showing how excess-zero probabilities interact with mean abundances using a simple bell-shaped mean shape (black line). The shape of the mean here doesn't really matter, it's more about looking at how excess probabilities decrease relative to mean abundance. The difference between zero-probabilities for the plain `nb` are strikingly lower than that of the `zinbl`, where `nb` zero-probability quickly reduces as the mean increases, while the `zinbl` model keeps it a lot higher till mean abundances almost reach their peak:

```{r check-prior-zero, fig.asp=0.3}
check_prior_zero(spec)
```

*OPTIONAL*: refine model spec further

## C. Fit subset model

Hopefully you're happy with the model specification and prior checks above? So we can proceed to some posterior checks using a "subset" mudel:

```{r fit-subset, results='hide'}
fit_subset <-
  mskt_fit(spec,
           use_subset = TRUE,
           iter_warmup = 200,
           iter_sampling = 100,
           chains = 6,
           parallel_chains = 6,
           # for debugging
           show_messages = TRUE, show_exceptions = TRUE)
```

A few divergences like this raise an initial warning that the posterior is not simple to compute. Summaries for the `m` parameter confirm this, where some of the chains may not have mixed (independently sampled the posterior parameter space):

```{r fit-subset-check}
check_computation(fit_subset)
```

*OPTIONAL*: refine model spec or fit parameters

### Check subset model

To dig deeper into the high `rhat`, $\hat{R}$, values for `m` we can look at the marginal posterior distributions for each chain:

```{r check-post-dens-subset, fig.asp=0.8}
# visual display of chain mixing and prior-data conflicts
check_post_dens(fit_subset, by_chain = TRUE)
```

The `m` issue isn't apparent visually, there's something with `m` and `r` discussed in thesis.

Moving forward, we can compare the CDF of the posterior predictive distribution vs the empirical CDF's of the training and test data using a discrete probability integral transform plot:

```{r check-post-calibration-subset}
# uses test set for discrete pit
check_post_calibration(fit_subset)
```

The little uptick at the end suggests that the posterior distribution predicts both training (subset, seen) data and test (unseen) data well up until the very highest counts, where it might be slightly underdispersed (i.e., underpredicting max counts). We could check this with the `bayesplot` package using similar methods as `check_prior_dist` above.

```{r}
bayesplot::ppc_stat(y = spec$subset()$train$y,
                    yrep = fit_subset$draws('y_rep', format = 'matrix'),
                    stat = 'max')
```

*OPTIONAL*: refine model spec or fit parameters

## D. Fit full model

Alright, some concerns with chains not mixing above, but not enough to suggest gross posterior misspecification. It could just be that, with the added complexity of the `zinbl` model and the sparser data, the computation needs a little more time to learn the posterior curvature. We do this by increasing the warm-up iterations to the default 1000 per chain:

```{r fit-full, results='hide'}
fit_full <-
  mskt_fit(spec,
           iter_warmup = 1000,
           iter_sampling = 1000,
           chains = 4,
           parallel_chains = 4,
           # we could also take smaller steps around the posterior using
           # adapt_delta > 0.8, but better to avoid if possible
           # for testing
           show_messages = TRUE, show_exceptions = TRUE)
```

Less divergences than before, increasing `adapt_delta` should clear those. The parameter diagnostics all look good:

```{r fit-full-check}
check_computation(fit_full,
                  # to save console space
                  hide_stats = c('mean', 'sd'))
```

*OPTIONAL*: refine model spec or fit parameters

### Check full model

Can check the marginal posterior distributions for each parameter again:

```{r}
check_post_dens(fit_full, by_chain = TRUE)
```

Much more certain, or precise. Using approximate leave-one-out cross-validation (LOO-CV) we can identify data points the posterior struggles to predict:

```{r check-post-influencers-full}
# can be slow for large N
check_post_influencers(fit_full)
```

A few zero-counts in mid-range temperatures, possibly the complex interaction between mean shape, dispersion, and linked zero-inflation struggling ever so slightly.

*OPTIONAL*: refine model spec or fit parameters.

## E. Use the model

To the fun(ner) part!

### Plot summaries of the abundance distribution

We'll plot the non zero-inflated `modskurt` mean here to get an idea of what average redstripe CPUE would be in the absence of excess-zeros:

```{r use-full-fit-dist}
# plot the distribution of abundance along the gradient
abundance_dist(fit_full,
               include_zero_inflation = FALSE,
               summaries = c('mean')) +
  scale_y_sqrt()
```

### Calculate ranges of x for different percentages of abundance measures

And a quick look at the temperatures that have greater than half of the highest predicted mean CPUE:

```{r use-full-fit-range-plot}
abundance_range(fit_full,
                capture_pct = 50,
                using_range = 'HAZ',
                based_on = 'mean',
                include_zero_inflation = FALSE,
                plotted = TRUE) +
  labs(subtitle = '50% of highest mean counts') +
  scale_y_sqrt()
```

About $6.3$ to $7.5$ degree celsius, this may be useful, somewhere?

## Summary

This article stepped through the `modskurt` workflow with a ZINBL (zero-inflated negative binomial with excess-zero probability linked to the negative binomial mean) model of redstripe rockfish abundance at different water temperatures. The ZINBL model is very powerful, but possibly limited to very sparse data and careful prior specification to reliably estimate a credible posterior over a more complex likelihood surface.
